Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists

### 2024-03-22 paper.Rmd
Updated paper based on feedback Karsten Lambers

---
title: "Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists"
author:
  - name: Ronald M. Visser
    orcid_id: 0000-0001-6966-1729
    roles: [writing, editing, analysis, visualization]
    affiliation: Saxion University of Applied Sciences
  - name: Annemarie Jutte
    orcid_id: 0009-0004-9322-0674
    roles: [writing, editing, software]
    affiliation: Saxion University of Applied Sciences
  - name: Doug Rocks-Macqueen 
    orcid_id: 0000
    roles: [writing, editing, software]
    affiliation: Landward Research Teoranta
  - name: Iza Romanowska 
    orcid_id: 0000-0002-9487-2111
    roles: [writing, editing]
    affiliation: Aarhus University
  - name: Tom Brughmans
    roles: [writing, editing]
    orcid_id: 0000-0002-1589-7768
    affiliation: Aarhus University
  - name: Laura van der Knaap 
    roles: [writing, editing, software]
    orcid_id: 0000
    affiliation: Leiden University
  - name: Karsten Lambers 
    roles: [writing, editing]
    orcid_id: 0000-0001-6432-0925
    affiliation: Leiden University
  - name: Kenneth Aitchison 
    roles: [writing, editing, software, project administration]
    orcid_id: 0000-0001-6594-1408
    affiliation: Landward Research Teoranta
  - name: Adam Pažout
    roles: [writing, editing]
    orcid_id: 0000
    affiliation: Aarhus University
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  distill::distill_article:
    number_sections: yes
    fig_caption: yes
  bookdown::pdf_document2:
    number_sections: yes
    keep_tex: true
    fig_caption: yes
    pandoc_args:
      --lua-filter=scholarly-metadata.lua
  bookdown::word_document2:
    number_sections: yes
    fig_caption: yes
bibliography: references.bib
csl: journal-of-computer-applications-in-archaeology.csl
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
  output_format = "all") })
---

@@

Other teaching material available SPOC [@scherjon2019]

@@

During the project we have developed tutorials. The tutorials were first designed using storyboards which were later converted into tutorials using NetLOGO Web (<https://www.netlogoweb.org/launch>) and Javascript (). The tutorials were developed together and we worked together using GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials>). During the project initial testing was done with all authors and issues in GitHub were used to tackle bugs.

The learning and teaching material for ABM for archaeologists assumes that the users have completed secondary education. It is suitable for archaeologists with at least some background in archaeology and preferably after at least the first year of a Bachelor study in Archaeology (preferably above EQF levels 4 or 5). In addition, the tutorials are aimed at professionals working in archaeology and some experience with computer. We also assume that learners are proficient in English and have at least the Reading B1 level, but Reading B2 is recommended [@councilofeurope2020]. The tutorials are developed aiming to develop learners in the following competence area's: 1. Information and data literacy, 2. Communication and collaboration, 3. Digital content creation and, 5. Problem solving [@europeancommissionjointresearchcentre2022].

The tutorials were tested during various online and in-person conferences. The in-person conferences were attended by several people from the project team. During the online events we were always present with several people and enabled participants to go to break-out rooms if they needed any help or wanted to discuss things. In addition, we also had students working with the tutorials at Saxion University of Applied Sciences and Leiden University. We used the feedback of the participants to improve the tutorials. The feedback was obtained structurally from two surveys using Qualtrics. The participants of the events were asked to answer questions of the first survey before participating. We used the answers to understand what the background of the participants was and to get an estimate of their level of knowledge in relation to ABM (see appendix ... for the questions). After the participants worked with the tutorials during the event they were asked to answer a second survey. Questions of this survey were aimed at the measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix ... for the questions). For some events participants had to register beforehand and we were able to send the pre-workshop survey by email. At other events no registration was possible or necessary and the pre-workshop survey was given at the start of the event. The post-workshop survey was distributed using QR-codes or links at the end of the workshop during the event. After each event we have improved the tutorials based on the received feedback.

| Conference or event          | Period                  | Online, in person or hybrid | Number of participants | Number of registrations |
|---------------|---------------|---------------|---------------|---------------|
| CAA Amsterdam                | April 2023              | In person                   | 36                     | 53                      |
| EAA Belfast                  | August 2023             | In person                   | 21                     | No registration         |
| CAA-DE/NL-Fl Online workshop | October 2023            | Online                      | 40                     | 76                      |
| Reuvensdagen                 | November 2023           | In person                   | 13                     | No registration         |
| CAA-UK                       | November 2023           | Hybrid                      | 53                     |                         |
| Leiden (course)              | December 2023           | In person                   | 11                     | No registration         |
| Aarhus                       | January/February 2024   | Online                      | 176                    | \>500                   |
| Saxion                       | March 2023-January 2024 | In person                   | 18                     | No registration         |
| **Total**                    |                         |                             | **368**                | **\>628**               |

: Conferences, events and situations were the workshops given and the tutorials were tested.

@@

Over the course of the project three groups of students worked on the project during the Smart Solutions Semester at Saxion University of Applied Sciences. This is an interdisciplinary semester in which students of at least three different study-programmes or disciplines work together on a complex problem/project (<https://www.saxion.edu/business-and-research/collaborate-with-saxion/smart-solutions>). The backgrounds of the various students were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media & Game Technologies, and ICT. The diverse groups of students contributed with new ideas, developing educational material, testing the tutorials and developing a style for the website and other materials.

@@

In the first tutorial the learner will learn what simulations and agent-based models are and how they can help in archaeological research. This tutorial consists of 4 lessons. The first two lesson introduce the learner into simulation in general and Agent-Based Modelling in specific. Various concepts related to ABM are introduced and the first concepts within NetLogo are explained. The third lesson aims to explain how ABMs are used in archaeological research. The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode. This tutorial build a foundation for working in NetLogo and with ABM. All activity stays within the intermediate level, although higher levels might be touch upon.

In tutorial 2 the learner will learn the basics of NetLogo by making your first simulation on the Out of Africal dispersal of homo sapiens. The leaerner will work with basic NetLogo syntax and learn how to set up a simulation and visualize their outcomes. This tutorial consists of 9 lessons. This tutorial intends to guide the learner from the intermediate to the advanced level of proficiency. The learning curve is relatively gentle. This is achieved by learning to work with the NetLogo web interface. The primitives are introduced and initialization phase of a simple ABM is gone through. The world within NetLogo's world (including dimensions, coordinates, origin) and how to alter it are explained and experience in a hands-on approach. The learners learn about simple simulation loops and how to use primitives. The advanced level is achieved with the custom procedures and variables. In addition, if-statements are introduced and more complex versions thereof. Exporting information using plots is also explained.

@@

In tutorial 4 the learner will work with Sugarscape simulations to further expand your ABM and NetLogo skills. The learner will learn how to set up more complex interactions between agents and the environment. Furthermore, the basics of setting up good experiments and validating models will be explained. This tutorial consists of 8 lessons. This tutorial intends to guide the learner from the (intermediate and) advanced level to the highly specialised level of proficiency. In this tutorial more complex agent-environment interaction is further learned, including learning ways to visualize the environment and how to give more agency to patches. Creating toy landscapes are central to lesson 5. The learner also learns the basics of setting up experiments in NetLogo and how to use monitors and flexible plots to understand the results. The validation of agent-based models is an important concept of designing a good experiment. The finale step, is to compare simulation results with the archaeological record. The last lesson is aimed at learning syntax to make lists more dynamic and how to refactor code.

In the final and fifth tutorial the learner will learn more about how to actually incorporate agent-based modelling in research. This tutorial focusses less on programming in NetLogo and more on the model development process. The learner will learn about the different phases in modelling and how to export data from models. This tutorial consists of 5 lessons. This tutorial touches upon various levels, but mainly between the intermediate and highly specialised level of proficiency. This tutorial is approaches more theoretical aspects in a practical environment. The learner leans about the model development process starting with the conceptual phase of the model development. Good and bad research questions for modelling are explained and the difference between them. The learner needs to understand how to pick the right modelling technique. The learner should understand the importance of properly conceptualizing a model before starting the technical phase of the model development. In this phase parametrisation, designing experiments and the analyses and interpretation of the models are important. Learning about the dissemination phase of the model development is aimed at understanding (the importance of) publishing models and replication. In addition, learning how to export results from NetLogo using basic export primitives is explained. The BehaviorSpace is explained to enable the learners to analyse the output of models.

@@

A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants). The respondents came from many different countries all over the world (see Figure \@ref(fig:nationality). The number of female respondents slightly outnumber the male ones and a small group did not share their gender and two were non-binary (see Figure \@ref(fig:gender-age)). The ages of the respondents ranged from below 20 to over 70, reaching people in different age classes, although the majority was aged between 20 and 40. It seems that the different events also had a slightly different distribution of both gender and age. For example, more older people attended the workshop at the CAA conference in April 2023 and more male people were present during the workshop at the Reuvensdagen in November 2023.

@@

```{r gender-age, fig.cap="The gender and age of the respondents for each workshop", fig.width=11, echo=FALSE}
ggplot(pre_surveys, aes(x = `What.is.your.age.`, fill = `What.is.your.gender....Selected.Choice`)) +
  geom_bar() + labs(x= "Age", y="Count", fill = "Gender") + facet_grid(~Workshop) +   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

@@

As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply this or had never applied this before. The respondents were also asked how they rated the available knowledge on ABM (Figure \@ref(fig:available-theory)). While the largest group of them had no opinion on the subject, a very large proportion of the responded answered that they rated the available theory on ABM as limited and a smaller group as sufficient. A minority rated the available theory on ABM as more than sufficient. This clearly shows the need for better educational material.

@@

During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials. Some particpants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general.

## After workshop

@@

Two more open questions were asked to the respondents. The first was aimed at getting positive feedback and the other was aimed at getting feedback to improve the tutorials.

@@

We also received feedback on possible improvements. For the first workshops we mostly received feedback on the number of bugs. During the CAA-conference in Amsterdam we did the first test of the tutorials and we also instructed the participants that this was work in progress. The number of references to bugs reduced to zero by the end and people started providing other feedback, for example the wish to see more examples or expanding the tutorials more. Some wanted more cooperation, which is possible, but for some online events harder, although we provided break-out rooms. It is also interesting to note that more and more people did not see any room for improvement.

The majority of the respondents that wanted to apply ABM in the future for research (144 of the 164 that answered this question, see Figure \@ref(fig:future-abm)). Some of the respondents were already using ABM. A large group was not sure yet in how to apply ABM and wanted to read more on the subject or play a bit with the possibilities. Many respondents also shared the context in which they were thinking to apply ABM to. Various were thinking about movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms. Others thought of demography, social networks, migration or settlement distributions patterns. The natural environment and the interaction with humans in the past was also mentioned by some, and often in relation to GIS or how to replace GIS with ABM. The archaeological periods that the participants were interested in were very diverse ranging from the paleolithic to the medieval period.

@@

The tutorials are aimed at developing learners in the following competence area's: 1. Information and data literacy, 2. Communication and collaboration, 3. Digital content creation and, 5. Problem solving [@europeancommissionjointresearchcentre2022]. In these competence area's various competences are developed towards an advanced or even specialised level.

@@

For Competence Area 2. Communication and collaboration the tutorials mainly address 2.1 Interacting Through Digital Technologies, because the users are constantly interacting with digital technologies, but communications is not relevant for the tutorials. However, during the workshops given in the course of the project, participants interacted with each other and the teachers.

Competence Area 3. Digital content creation is one of the main competence area's for the ABM-teaching material. In the process of learning ABM the users are constantly 3.1 Developing Digital Content and 3.2 Integrating And Re-Elaborating Digital Content. The material does not touch upon 3.3 Copyright And Licences, but we assume that users have a general understanding. A very important aspect of learning ABM is 3.4 Programming. The users are exploring the possibilities and chances of programming an ABM. The users are expected to learn both syntax and more general concepts such as modular code development, loops, lists and commenting and documenting code.

Competence Area 5. Problem solving is very important and is closely tied to 3.4 Programming, since programming involves a lot of problem solving. In addition, the creation of a highly complex ABM consists of problem solving all the time, including debugging. All competencies are addressed: 5.1 Solving Technical Problems, is common when writing code, learning about coding, error handling and going from pseudocode to real code. The competence 5.2 Identifying Needs And Technological Responses, is a central issue of the course, while 5.3 Creatively Using Digital Technology is an important aspect of learning to think in models and develop models themselves. In tutorial 2 various aspects of 5.4 Identifying Digital Competence Gaps, are relevant, since this tutorial will force the user to remember code and commands and forces the user to assess their knowledge and skills.

>>>>

---
title: "Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists"
author:
  - name: Ronald M. Visser
    orcid_id: 0000-0001-6966-1729
    roles: [writing, editing, analysis, visualization]
    affiliation: Saxion University of Applied Sciences
  - name: Annemarie Jutte
    orcid_id: 0009-0004-9322-0674
    roles: [writing, editing, software]
    affiliation: Saxion University of Applied Sciences
  - name: Doug Rocks-Macqueen 
    orcid_id: 0000
    roles: [writing, editing, software]
    affiliation: Landward Research Teoranta
  - name: Iza Romanowska 
    orcid_id: 0000-0002-9487-2111
    roles: [writing, editing]
    affiliation: Aarhus University
  - name: Tom Brughmans
    roles: [writing, editing]
    orcid_id: 0000-0002-1589-7768
    affiliation: Aarhus University
  - name: Laura van der Knaap 
    roles: [writing, editing, software]
    orcid_id: 0000
    affiliation: Leiden University
  - name: Karsten Lambers 
    roles: [writing, editing]
    orcid_id: 0000-0001-6432-0925
    affiliation: Leiden University
  - name: Kenneth Aitchison 
    roles: [writing, editing, software, project administration]
    orcid_id: 0000-0001-6594-1408
    affiliation: Landward Research Teoranta
  - name: Adam Pažout
    roles: [writing, editing]
    orcid_id: 0000
    affiliation: Aarhus University
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  distill::distill_article:
    number_sections: yes
    fig_caption: yes
  bookdown::pdf_document2:
    number_sections: yes
    keep_tex: true
    fig_caption: yes
  bookdown::word_document2:
    number_sections: yes
    fig_caption: yes
bibliography: references.bib
csl: journal-of-computer-applications-in-archaeology.csl
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
  output_format = "all") })
---

@@

We were able to draw on previous experiences with a small private online course (SPOC) on 'Modelling and Simulation in Archaeology' that some of us designed for graduate teaching at the Faculty of Archaeology, Leiden University, where we taught this course in 2016/17 and again in 2018/19 [@scherjon2019]. For this course we developed new online teaching materials comprising pre-recorded lectures, practical exercises, reading assignments and exams. While we greatly benefited from that experience, the target group of the SPOC was graduate students with prior experience in digital archaeology, whereas the ERASMUS+ project was aimed at a much broader target group of archaeology students and professionals with only basic digital skills.

@@

During the project we developed online tutorials. The tutorials were first designed using storyboards which were later converted into tutorials using NetLOGO Web (<https://www.netlogoweb.org/launch>) and Javascript (). The tutorials were developed as a team and we worked together using GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials>). During the project, initial testing was done by all authors and issues in GitHub were used to tackle bugs.

The learning and teaching material for ABM for archaeologists assumes that the users have completed secondary education. It is suitable for archaeologists with at least some background in archaeology, preferably after the first year of a Bachelor study in Archaeology (above EQF levels 4 or 5). In addition, the tutorials are aimed at professionals working in archaeology who have some experience with computer applications. We also assume that learners are proficient in English and have at least the Reading B1 level, but Reading B2 is recommended [@councilofeurope2020]. The tutorials are developed aiming to develop learners' skills in the following competence area's: 1. Information and data literacy, 2. Communication and collaboration, 3. Digital content creation and, 5. Problem solving [@europeancommissionjointresearchcentre2022].

After initial debugging by the project team, the tutorials were tested 'in the wild' during various online and in-person conferences. The in-person conferences were attended by several people from the project team. During the online events we were always present with several people and enabled participants to go to break-out rooms if they needed any help or wanted to discuss things. In addition, we also had students working with the tutorials at Saxion University of Applied Sciences and Leiden University. We We obtained structured feedback from the participants and used it to improve the tutorials prior to the next test. The feedback was obtained from two surveys using Qualtrics. The participants of the events were asked to answer questions of the first survey before participating. We used the answers to understand what the background of the participants was and to get an estimate of their level of knowledge in relation to ABM (see appendix ... for the questions). At the end of the event the participants were asked to answer a second survey. Questions of this survey were aimed at measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix ... for the questions). For some events participants had to register beforehand and we were able to send the pre-workshop survey by email. At other events no registration was possible or necessary and the pre-workshop survey was given at the start of the event. The post-workshop survey was distributed using QR-codes or links at the end of the workshop during the event. After each event we improved the tutorials based on the received feedback.

| Conference or event          | Period                  | Online, in person or hybrid | Number of participants | Number of registrations |
|------------------------------|-------------------------|-----------------------------|------------------------|-------------------------|
| CAA Amsterdam                | April 2023              | In person                   | 36                     | 53                      |
| EAA Belfast                  | August 2023             | In person                   | 21                     | No registration         |
| CAA-DE/NL-Fl Online workshop | October 2023            | Online                      | 40                     | 76                      |
| Reuvensdagen                 | November 2023           | In person                   | 13                     | No registration         |
| CAA-UK                       | November 2023           | Hybrid                      | 53                     |                         |
| Leiden (course)              | December 2023           | In person                   | 11                     | No registration         |
| Aarhus                       | January/February 2024   | Online                      | 176                    | \>500                   |
| Saxion                       | March 2023-January 2024 | In person                   | 18                     | No registration         |
| **Total**                    |                         |                             | **368**                | **\>628**               |

: Conferences, events and situations were workshops were held at which the tutorials were tested.

@@

Over the course of the project three groups of students worked on the project during the Smart Solutions Semester at Saxion University of Applied Sciences. This is an interdisciplinary semester in which students of at least three different study-programmes or disciplines work together on a complex problem/project (<https://www.saxion.edu/business-and-research/collaborate-with-saxion/smart-solutions>). The backgrounds of the various students were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media & Game Technologies, and ICT. The diverse groups of students contributed new ideas, developing educational material, testing the tutorials and developing a style for the website and other materials.

@@

In the first tutorial the user will learn what simulations and agent-based models are and how they can help in archaeological research. This tutorial consists of 4 lessons. The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific. Various concepts related to ABM are introduced and the first concepts within NetLogo are explained. The third lesson aims to explain how ABMs are used in archaeological research. The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode. This tutorial build a foundation for working in NetLogo and with ABM. All activity stays within the intermediate level, although higher levels might be touched upon.

In tutorial 2 the user will learn the basics of NetLogo by making your their simulation on the Out of Africal dispersal of homo sapiens. The learner will work with basic NetLogo syntax and learn how to set up a simulation and visualize its outcomes. This tutorial consists of 9 lessons. This tutorial intends to guide the learner from the intermediate to the advanced level of proficiency. The learning curve is relatively gentle. This is achieved by learning to work with the NetLogo web interface. The primitives are introduced and initialization phase of a simple ABM is gone through. The world within NetLogo's world (including dimensions, coordinates, origin) and how to alter it are explained and experience in a hands-on approach. The learners learn about simple simulation loops and how to use primitives. The advanced level is achieved with the custom procedures and variables. In addition, if-statements are introduced and more complex versions thereof. Exporting information using plots is also explained.

@@

In tutorial 4 the learner will work with Sugarscape simulations to further expand their ABM and NetLogo skills. The learner will learn how to set up more complex interactions between agents and the environment. Furthermore, the basics of setting up good experiments and validating models will be explained. This tutorial consists of 8 lessons. This tutorial intends to guide the learner from the (intermediate and) advanced level to the highly specialised level of proficiency. In this tutorial more complex agent-environment interaction is further learned, including ways to visualize the environment and how to give more agency to patches. Creating toy landscapes is central to lesson 5. The learner also learns the basics of setting up experiments in NetLogo and how to use monitors and flexible plots to understand the results. The validation of agent-based models is an important concept of designing a good experiment. The finale step, is to compare simulation results with the archaeological record. The last lesson is aimed at learning syntax to make lists more dynamic and how to refactor code.

In the final and fifth tutorial the learner will learn more about how to actually incorporate agent-based modelling in research. This tutorial focusses less on programming in NetLogo and more on the model development process. The learner will learn about the different phases in modelling and how to export data from models. This tutorial consists of 5 lessons. This tutorial touches upon various levels, but mainly between the intermediate and highly specialised level of proficiency. This tutorial approaches more theoretical aspects in a practical environment. The learner leans about the model development process starting with the conceptual phase of the model development. Good and bad research questions for modelling are explained and the difference between them. The learner needs to understand how to pick the right modelling technique. The learner should understand the importance of properly conceptualizing a model before starting the technical phase of the model development. In this phase parametrisation, designing experiments and the analysis and interpretation of the models are important. Learning about the dissemination phase of the model development is aimed at understanding (the importance of) publishing models and replication. In addition, learning how to export results from NetLogo using basic export primitives is explained. The BehaviorSpace is explained to enable the learners to analyse the output of models.

@@

A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants). The respondents came from many different countries all over the world (see Figure \@ref(fig:nationality). The number of female respondents slightly outnumbers the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \@ref(fig:gender-age)). The ages of the respondents ranged from below 20 to over 70, reaching people in different age classes, although the majority was aged between 20 and 40. It seems that the different events also had a slightly different distribution of both gender and age. For example, more older people attended the workshop at the CAA conference in April 2023 and more male people were present during the workshop at the Reuvensdagen in November 2023.

@@

```{r gender-age, fig.cap="The gender and age distribution of the respondents for each workshop", fig.width=11, echo=FALSE}
ggplot(pre_surveys, aes(x = `What.is.your.age.`, fill = `What.is.your.gender....Selected.Choice`)) +
  geom_bar() + labs(x= "Age", y="Count", fill = "Gender") + facet_grid(~Workshop) +   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

@@

As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply it or had never applied it before. The respondents were also asked how they rated the available knowledge on ABM (Figure \@ref(fig:available-theory)). While the largest group of them had no opinion on the subject, a very large proportion of the responded answered that they rated the available theory on ABM as limited and a smaller group as sufficient. A minority rated the available theory on ABM as more than sufficient. This clearly shows the need for better educational material.

@@

During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials. Some particpants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general. Individual progress varied depending on skills and preferences of the participants. While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later instalments.

## After the workshop

@@

Two more open questions were asked to the respondents. The first was aimed at learning what aspects the participants liked in the tutorials, and the other was aimed at getting feedback to improve the tutorials.

@@

We also received feedback on possible improvements. For the first workshops we mostly received feedback on the number of bugs. During the CAA-conference in Amsterdam we did the first test of the tutorials and we also instructed the participants that this was work in progress. The number of references to bugs reduced to zero by the end and people started providing other feedback, for example the wish to see more examples or expanding the tutorials more. Some wanted more cooperation, which is possible, but harder for some online events, although we provided break-out rooms. It is also interesting to note that over time more and more people did not see any room for improvement.

The majority of the respondents wanted to apply ABM in the future for research (144 of the 164 that answered this question, see Figure \@ref(fig:future-abm)). Some of the respondents were already using ABM. A large group was not sure yet how to apply ABM and wanted to read more on the subject or play a bit with the possibilities. Many respondents also shared the context to which they were thinking to apply ABM. Various participants mentioned movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms. Others thought of demography, social networks, migration or settlement distributions patterns. The natural environment and the interaction with humans in the past was also mentioned by some, and often in relation to GIS or how to replace GIS with ABM. The archaeological periods that the participants were interested in were very diverse, ranging from the Paleolithic to the Medieval period.

@@

As mentioned above, the following competence area's were identified as relevant for the tutorials: 1. Information and data literacy, 2. Communication and collaboration, 3. Digital content creation and, 5. Problem solving [@europeancommissionjointresearchcentre2022]. In these competence area's various competences are developed towards an advanced or even specialised level.

@@

For Competence Area 2. Communication and collaboration the tutorials mainly address 2.1 Interacting Through Digital Technologies, because the users are constantly interacting with digital technologies, but communication is not relevant for the tutorials. However, during the workshops given in the course of the project, participants interacted with each other and the teachers.

Competence Area 3. Digital content creation is one of the main competence areas for the ABM-teaching material. In the process of learning ABM the users are constantly 3.1 Developing Digital Content and 3.2 Integrating And Re-Elaborating Digital Content. The material does not touch upon 3.3 Copyright And Licences, but we assume that users have a general understanding. A very important aspect of learning ABM is 3.4 Programming. The users are exploring the possibilities and chances of programming an ABM. The users are expected to learn both syntax and more general concepts such as modular code development, loops, lists and commenting and documenting code.

Competence Area 5. Problem solving is very important and is closely tied to 3.4 Programming, since programming involves a lot of problem solving. In addition, the creation of a highly complex ABM consists of problem solving all the time, including debugging. All competencies are addressed: 5.1 Solving Technical Problems, is common when writing code, learning about coding, error handling and going from pseudocode to real code. The competence 5.2 Identifying Needs And Technological Responses, is a central issue of the course, while 5.3 Creatively Using Digital Technology is an important aspect of learning to think in models and develop models themselves. In tutorial 2 various aspects of 5.4 Identifying Digital Competence Gaps, are relevant, since this tutorial will force the user to remember code and commands and to assess their knowledge and skills.
